---
title: "Starting with data"
author: "Data Carpentry contributors"
minutes: 20
---

```{r, echo=FALSE, purl=FALSE, message = FALSE}
source("setup.R")
```

------------

> ### Learning Objectives
>
> * Load external data from a .csv file into a data frame.
> * Describe what a data frame is.
> * Summarize the contents of a data frame.
> * Use indexing to subset specific portions of data frames. 


------------

## Our question: which genes are changing?

This workshop is based around a common situation in transcriptomics. We are interested in a group of genes, and another study has measured expression in  conditions where expression these genes might be interesting. Can we tell if our favourite genes 

Here the dataset is of cell cycle progression in the yeast *Saccharomyces cerevisiae*: Translational control of lipogenic enzymes in the cell cycle of synchronous, growing yeast cells. Blank et al 2017 https://doi.org/10.15252/embj.201695050.


In figure 2, the paper notes that ribosome biogenesis genes are periodically expressed during the cell cycle, but doesn't tell us which ones. Can we find if our favourite gene is indeed periodically expressed? So our data analysis goals are:

* Find which ribosome biogenesis genes are on the list in figure 2
* Plot the individual trajectories of favourites: NOP56, ACT1
* Recreate figure 2 so we can read the gene names.

```{r, echo=FALSE, purl=TRUE}
### Presentation of the survey data
```

### Which ribosome biogenesis genes are in the list?

We need a list of ribosomal biogenesis (ribi) genes, and the list from figure 2 of the paper.

#### List of ribosomal biogenesis (ribi) genes.

For this lesson, we will use the list of yeast ribi genes from https://www.yeastgenome.org/go/GO:0042254

Go to the *bottom* of the page for the longer computational list, click download and place the file in your `data` directory. You'll need to have a folder on your machine called "data" where 
you'll download the file. Take a look at it: this file is a tab-delimited text file. 

We are going to use the tidyverse function `read_tsv()` to read the contents of this file into R as a tibble (nice data frame) object. Inside the `read_tsv()` command, the first entry is a character string with the file name ("data/ribosome_biogenesis_annotations.txt")

```{r, eval=TRUE, purl=TRUE}
read_tsv("data/ribosome_biogenesis_annotations.txt")
```

What does this tell us about the file?

Because we will need to reuse, let's make the list into an object

```{r, eval=TRUE,  purl=FALSE}
ribi_annotation <- read_tsv("Downloads/ribosome_biogenesis_annotations.txt")
```

This statement doesn't produce any output because, as you might recall,
assignments don't display anything. If we want to check that our data has been
loaded, we can see the contents of the data frame by typing its name: `ribi_annotation`.

```{r, eval = FALSE, purl = FALSE}
## Try also
View(ribi_annotation)
```
